---
title: "NYPD_Shooting_Incident_Data"
author: "Sashaank"
date: "2025-03-17"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)
```

```{r}
#load csv files for easy reproducibility
url_NYPD <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
NYPD <- read.csv(url_NYPD)
#column names of NYPD
colnames(NYPD)
#structure of data set
str(NYPD)
#summary of the raw data set from the website
summary(NYPD)

```
```{r }
NYPD_clean <- NYPD %>%
  #selecting relevant columns for data analysis
  select(OCCUR_DATE, OCCUR_TIME, BORO, PRECINCT, STATISTICAL_MURDER_FLAG, 
         VIC_AGE_GROUP, VIC_SEX, VIC_RACE) %>%
  mutate(
    OCCUR_DATE = mdy(OCCUR_DATE), #converting date from chr
    OCCUR_TIME = hms(OCCUR_TIME), #converting time from chr
    # Convert categorical variables
    VIC_AGE_GROUP = as.factor(VIC_AGE_GROUP),
    VIC_SEX = as.factor(VIC_SEX),
    VIC_RACE = as.factor(VIC_RACE),
    BORO = as.factor(BORO),
    STATISTICAL_MURDER_FLAG = as.logical(STATISTICAL_MURDER_FLAG), #true/false
    Year = year(OCCUR_DATE), #extract year - helps with analysis of trends of time (over years)
    Month = month(OCCUR_DATE, label = TRUE), #extract month (seasonal analysis? - are there more crimes during a certain season?)
    Day = day(OCCUR_DATE), #extract day - do shootings occur during certain days more?
    Hour = hour(OCCUR_TIME), #extract time - at specific times of the day?
    Shootings = 1 #count
  )
#data structure after cleaning the dataset
str(NYPD_clean)
#summary of the dataset
summary(NYPD_clean)

```

```{r}
NYPD_clean %>%
  ggplot(aes(x = BORO, fill = BORO)) + #fill makes it so each bar is a different color otherwise all would be same color
  geom_bar() + #adds a bar chart
  labs( #labels
    title = "NYPD Shootings Incidents by Borough",
    subtitle = "(From 2006 - To 2021)",
    x = "NYC Boroughs",
    y = "Total Number of Shootings",
  ) +
  theme_minimal() #removes background grid / makes it easier to read 

```

```{r}
NYPD_hour <- NYPD_clean %>%
  group_by(Hour) %>%
  summarize(Total_Shootings = sum(Shootings))

# Plot
ggplot(NYPD_hour, aes(x = Hour, y = Total_Shootings)) +
  geom_line(color = "red", size = 1) +  # Line plot to show hourly trends
  geom_point(color = "blue", size = 3) +  # Points to mark exact values
  labs(
    title = "Shootings by Time of Day",
    subtitle = "Hourly Distribution of Shootings (2006-2021)",
    x = "Hour of the Day (24-hour)",
    y = "Total Shootings",
  ) +
  theme_minimal()
```

```{r}

NYPD_year <- NYPD_clean %>%
  group_by(Year) %>% #grouping data by year col
  summarize(Shootings = sum(Shootings)) #adds up all shootings (counts) for year 
  ungroup(NYPD_year)

NYPD_year %>%
  ggplot(aes(x = as.factor(Year), y = Shootings)) + #keep year as x axis, y as shootings since we're not using geom_bar now
  geom_point(color = "blue") + #color chosesn as blue for geom_point since it's a scatter point chart
  labs( #labels
    title = "NYPD Shooting Incidents by Year",
    x = "Year",
    y = "Total Shootings",
  ) +
  theme_minimal() #removes background grid / makes it easier to read 

```





```{r}
NYPD_boro_year <- NYPD_clean %>%
  group_by(BORO, Year) %>% #grouping by first borough then year
  summarize(Shootings = sum(Shootings)) #adds up all shootings/counts for each {bourough & year group}
  ungroup(NYPD_boro_year) #ungroup data after summarizing as this good practice

NYPD_boro_year %>%
  #color = BORO --> so we can see each total per borough per year
  ggplot(aes(x = Year, y = Shootings, color = BORO)) + #plotting with x-axis = year(has boroughs) and y-axis = shootings/counts
  geom_line() + #connects data points (line plot)
  geom_point() + #adds individual points like a scatter chart
  labs( #labels
    title = "NYPD Shootings by Borough Over Time",
    subtitle = "(2006 - 2021)",
    x = "Year",
    y = "Number of Shootings",
  ) +
  theme_minimal() #removes background grid / makes it easier to read 

```

In conclusion, the analysis of the NYPD shooting incidents from 2006 to 2021 provides some key insights into the way the shootings across different boroughs and years have changed. Brooklyn and the Bronx consistently had the highest number of incidents - suggesting a higher rate of crime whereas Staten Island had the least. The total number of shootings also fluctuated over the years and was noticeable that during the COVID era - that the number of shootings across the board was much lower in each borough. 
A potential source of bias could be that the dataset relies on NYPD reportings and not other unrepresented parties such as hospital records etc. COVID-19 lockdowns could have influenced the trends. The dataset does not also account for population changes in each of the boroughs which could also have an impact the true shooting numbers across the board.
